Energy stable discontinuous Galerkin formulations using discrete differential operators

### 2017-05-11 docs/splitform/main.tex
updating splitform docs and code with curvilnear

\cite{feng2016discontinuous}

@@

@@

We will assume for now that all elements are affine mappings of a reference element $\widehat{D}$.

@@

In order to define discrete differential operators, we first introduce the $L^2$ inner product on $V_h\LRp{\Oh}$
\[
\LRp{u,v}_{\Omega} = \sum_{D^k} \int_{D^k} uv \diff{x} = \sum_{D^k} \int_{\widehat{D}} uv J \diff{\widehat{x}},
\]
and the associated $L^2$ projection $\Pi_N: L^2\LRp{\Omega}\rightarrow P^N$
\[
\LRp{\Pi_N u,v}_{\Omega} = \LRp{u,v}_{\Omega}.  
\]

@@

\subsection{Discrete derivatives}
We introduce two discrete DG derivatives in this section.  The first is the implicit definition used in \cite{di2011mathematical}:
\begin{definition}
The discontinuous Galerkin differentiation operator ${D}^i: L^2\LRp{\Omega}\rightarrow V_h$ is defined implicitly as follows:
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{-u,\pd{v}{\bm{x}_i}}_{\LK} + \LRa{\avg{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v \in V_h.
\]
\end{definition}

We note that, for $u,v \in V_h$, (discrete) integration by parts yields an equivalent definition of $D^i$ found in \cite{hesthaven2004high,Warburton20063205}
\[
\LRp{{D}^i u,v}_{\Oh} = \sum_{D^k} \LRp{\LRp{\pd{u}{\bm{x}_i},v}_{\LK} - \frac{1}{2}\LRa{\jump{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v\in V_h.
\]
The main property of the discrete derivatives utilized in this work is a discrete integration-by-parts formula.  
\begin{lemma}
Let $u,v\in V_h$.  Then, 
\begin{align*}
\LRp{D^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i\\
\LRp{D_h^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
\end{align*}
\label{lemma:ibp}
\end{lemma}
\begin{proof}
Ez
\end{proof}

@@

Using these discrete derivatives, we can define discrete gradient and divergence operators
\begin{definition}
Let $u \in V_h$ and $\bm{u}\in \LRp{V_h}^d$.  Then, discrete gradient and divergence operators are defined as follows:
\begin{align*}
\LRp{\Grad^{\rm DG} u}_i &= D^i u, \qquad \Grad^{\rm DG}\cdot \bm{u} = \sum_{i=1}^d D^i \bm{u}_i\\
\LRp{\Grad^{\rm DG}_h u}_i &= D_h^i u, \qquad \Grad^{\rm DG}_h\cdot \bm{u} = \sum_{i=1}^d D_h^i \bm{u}_i.
\end{align*}
\end{definition}

\subsection{Examples}

\begin{itemize}
\item Tensor product elements: inner product computed using GLL nodes, $\Pi_N$ reduces to nodal interpolation
\item General elements: $\Pi_N$ is just quadrature-based projection.  For accuracy, degree $2N$ or higher?  
\item General elements: $\Pi_N$ can be taken to be nodal interpolation.  
\item Pyramids?  
\end{itemize}

\section{Notes}
Begin with the gradient.  Multiplication by $\bm{v}$ and integration by parts gives
\[
\sum_{D^k} \LRp{\Grad u, \bm{v}}_{\LK} = \sum_{D^k} \LRp{-u, \Grad \cdot \bm{v}}_{\LK} + \LRa{u,\bm{v}\cdot\bm{n}}_{\LdK}.
\]
We replace the values of ${u}$ with $\avg{u}$ on each element boundary to define the global DG gradient operator $\Grad_h$ 
\[
\sum_{D^k} \LRp{\Grad_h u, \bm{v}}_{\LK} \coloneqq \sum_{D^k} \LRp{-u, \Grad \cdot \bm{v}}_{\LK} + \LRa{\avg{u},\bm{v}\cdot\bm{n}}_{\LdK}.
\]
Integrating by parts again and the introduction of the lift operator shows that
\[
\Grad_h u = \Grad u + \frac{1}{2}L\LRp{\jump{u}\bm{n}}
\]
where $L$ is the lift operator.

The DG divergence operator is similarly defined as
\[
\sum_{D^k} \LRp{\Grad_h \cdot \bm{u}, v}_{\LK} \coloneqq \sum_{D^k} \LRp{-\bm{u}, \Grad v}_{\LK} + \LRa{\avg{\bm{u}}\cdot\bm{n},v}_{\LdK}.
\]
and
\[
\Grad_h \cdot u = \Grad\cdot u + \frac{1}{2}L\LRp{\jump{\bm{u}}\cdot\bm{n}}
\]
and it can be shown that 
\[
\LRp{\Grad_h u, \bm{v}} = \LRp{-u, \Grad_h \cdot \bm{v}}.
\]
We will incorporate boundary conditions in a stable way in the following sections.

@@

@@

It's probably better to formulate this using continuous DG derivatives, recover flux terms, then discretize that - the flux terms \textit{should} still cancel out after discretization, right?

\section{Variable advection}

The discrete DG derivative-based method is not consistent in the sense that Galerkin orthogonality does not hold exactly.  The difference lies in the flux terms.  Assume $\Div \bm{\beta} = 0$, then
\[
\LRp{\pd{u}{t},v} + \frac{1}{2}\LRp{-{ \bm{\beta}u},\Grad_h v} + \frac{1}{2}\LRp{\bm{\beta}\cdot\Grad_h u,v} = 0.
\]
\begin{align*}
\LRp{\Grad_h\cdot \Pi_N\LRp{ \bm{\beta}u},v} &= \sum_{D^k} \LRp{-\bm{\beta}u, \Grad v}_{\LK} + \LRa{\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n},v}_{\LdK}\\
\LRp{\Grad_h u,\bm{\beta}v} &= \sum_{D^k} \LRp{-u, \Grad \cdot \LRp{\bm{\beta}v}}_{\LK} + \LRa{\bm{\beta}\cdot\bm{n}\avg{u},v}_{\LdK}.
\end{align*}
The latter term is consistent; the former is not due to the presence of $\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n}$ in the flux term.  The consistency error should then be $O(h^{N+1/2})$ using a trace inequality for $L^2$ projections.

\section{Curvilinear coordinates}

For curvilinear meshes, $J$ and $\bm{G}$ are no longer constant over each element.  However, for isoparametric mappings (and special polynomial interpolations of general curvilinear mappings), metric identities hold \cite{kopriva2006metric} such that $\widehat{\Grad} \cdot \LRp{J\bm{G}} = 0$.  
\[
\LRp{\Grad u,\bm{v}}_{D^k} = \LRp{J\bm{G}\widehat{\Grad} u,\bm{v}}_{\widehat{D}} = \LRp{-u,\widehat{\Grad} \cdot \LRp{J\bm{G}\bm{v}}}  + \LRa{J\bm{G}\widehat{\bm{n}} u,\bm{v}} 
\]
Noting that $J\bm{G}\widehat{\bm{n}} = \bm{n} J^f$ \cite{hesthaven2007nodal}, we recover integration by parts over the physical element.

>>>>

Non-conservative discretizations \cite{rhebergen2008discontinuous}.

DG calculus \cite{feng2016discontinuous} usually assumes exact integration.  As pointed out in \cite{gassner2016split}, this can be either be very expensive (curvilinear meshes or polynomial nonlinearities) or impossible (in the presence of rational integrands, such as for compressible flow).

@@

Goal: reduce aliasing in advective problems caused by variable coefficients and curvilinear or non-affine mappings.

@@

We will assume for now that all elements are affine mappings of a reference element $\widehat{D}$.  Extensions to curvilinear meshes and non-affine mappings will be discussed in Section~\ref{sec:curv}.

@@

In order to define discrete differential operators, we first introduce the $L^2$ inner product on $V_h\LRp{\Oh}$
\[
\LRp{u,v}_{\Omega} = \sum_{k} \int_{D^k} uv \diff{x} = \sum_{k} \int_{\widehat{D}} uv J \diff{\widehat{x}},
\]
and the associated $L^2$ projection $\Pi_N: L^2\LRp{\Omega}\rightarrow P^N$
\[
\LRp{\Pi_N u,v}_{\Omega} = \LRp{u,v}_{\Omega}.  
\]

@@

\subsection{Discrete derivatives}
We introduce two discrete DG derivatives in this section, based on definitions used in \cite{di2011mathematical} and \cite{hesthaven2004high,Warburton20063205}:
\begin{definition}
We define two discontinuous Galerkin differentiation operators ${D}^i: L^2\LRp{\Omega}\rightarrow V_h$ and $\tilde{D}^i: L^2\LRp{\Omega}\rightarrow V_h$ is defined implicitly as follows:
\begin{align}
\LRp{{D}^i u,v}_{\Oh} &= \sum_{k} \LRp{\LRp{-u,\pd{v}{\bm{x}_i}}_{\LK} + \LRa{\avg{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v \in V_h\\
\LRp{\tilde{D}^i u,v}_{\Oh} &= \sum_{k} \LRp{\LRp{\pd{u}{\bm{x}_i},v}_{\LK} - \frac{1}{2}\LRa{\jump{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v\in V_h.
\label{eq:dgd1}
\end{align}
\end{definition}

We note that, assuming $u,v \in V_h$ and sufficiently accurate quadrature, $D^i$ and $\tilde{D}^i$ are equivalent.  The main property of the discrete derivatives utilized in this work is a discrete integration-by-parts formula.  
\begin{lemma}
Let $u,v\in V_h$.  Then, 
\begin{align*}
\LRp{D^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
%\LRp{D_h^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
\end{align*}
\label{lemma:ibp}
\end{lemma}
\begin{proof}
Assuming that quadrature is sufficiently accurate for $u, v\in V_h$, integration by parts gives the equivalence. 
\end{proof}

@@

Using these discrete derivatives, we can define discrete gradient and divergence operators
\begin{definition}
Let $u \in V_h$ and $\bm{u}\in \LRp{V_h}^d$.  Then, discrete gradient and divergence operators are defined as
\begin{align*}
\LRp{\Grad^{\rm DG} u}_i &= D^i u, \qquad \Grad^{\rm DG}\cdot \bm{u} = \sum_{i=1}^d D^i \bm{u}_i\\
\LRp{\Grad^{\rm DG}_h u}_i &= D_h^i u, \qquad \Grad^{\rm DG}_h\cdot \bm{u} = \sum_{i=1}^d D_h^i \bm{u}_i.
\end{align*}
\end{definition}

@@

\section{Examples}

\begin{itemize}
\item Tensor product elements: inner product computed using GLL nodes, $\Pi_N$ reduces to nodal interpolation
\item General elements: $\Pi_N$ is just quadrature-based projection.  For accuracy, degree $2N$ or higher?  
\item Triangles and tetrahedra: lumped quadrature nodes.  Implies that mass lumping on triangles requires sufficiently accurate quadrature such that IBP holds.  This conclusion was first reached in the study of multi-dimensional SBP operators \cite{hicken2016multidimensional}.  
\item General elements: $\Pi_N$ can be taken to be nodal interpolation.  
\item Pyramids: need to account for non-affine mappings.  
\end{itemize}

\subsection{Variable advection}

The discrete DG derivative-based method is not consistent in the sense that Galerkin orthogonality does not hold exactly.  The difference lies in the flux terms.  Assume $\Div \bm{\beta} = 0$, then
\[
\LRp{\pd{u}{t},v} + \frac{1}{2}\LRp{-{ \bm{\beta}u},\Grad_h v} + \frac{1}{2}\LRp{\bm{\beta}\cdot\Grad_h u,v} = 0.
\]
\begin{align*}
\LRp{\Grad_h\cdot \Pi_N\LRp{ \bm{\beta}u},v} &= \sum_{k} \LRp{-\bm{\beta}u, \Grad v}_{\LK} + \LRa{\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n},v}_{\LdK}\\
\LRp{\Grad_h u,\bm{\beta}v} &= \sum_{k} \LRp{-u, \Grad \cdot \LRp{\bm{\beta}v}}_{\LK} + \LRa{\bm{\beta}\cdot\bm{n}\avg{u},v}_{\LdK}.
\end{align*}
The latter term is consistent; the former is not due to the presence of $\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n}$ in the flux term.  The consistency error should then be $O(h^{N+1/2})$ using a trace inequality for $L^2$ projections.

@@

\section{Non-affine mappings}
\label{sec:curv}

For general elements and curvilinear meshes, $J$ and $\bm{G}$ are no longer constant over each element.  However, for isoparametric mappings (and appropriate polynomial interpolations of general curvilinear mappings), metric identities hold \cite{kopriva2006metric} such that $\widehat{\Grad} \cdot \LRp{J\bm{G}^T} = 0$.  Thus, we can write
\[
\LRp{\Grad u,\bm{v}}_{D^k} = \LRp{J\bm{G}\widehat{\Grad} u,\bm{v}}_{\widehat{D}} = \LRp{-u,\widehat{\Grad} \cdot \LRp{J\bm{G}\bm{v}}}  + \LRa{J\bm{G}^T\widehat{\bm{n}} u,\bm{v}}.   
\]
Noting that $J\bm{G}\widehat{\bm{n}} = J^f \bm{n} $ \cite{hesthaven2007nodal}, we recover integration by parts over the physical element.

Thus, to extend the discrete discontinuous Galerkin gradient and divergence to non-affine mappings, we simply define
\begin{align*}
\LRp{\widehat{\Grad}_h u,\bm{v}}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}u,\bm{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{u},\bm{v}} \\
\LRp{\widehat{\Grad}_h \cdot\bm{u},{v}}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}\cdot \bm{u},{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\jump{\bm{u}},v\widehat{\bm{n}}} 
%\LRp{\Grad_h \cdot \bm{u},v}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}\cdot{\Pi_N\LRp{J\bm{G}\bm{u}}},v}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{J\bm{G}\bm{u}},v} \\
%\LRp{\Grad_h u,\bm{v}}_{\Omega} &= \sum_{k} \LRp{\Pi_N\LRp{J\bm{G}\widehat{\Grad} u},\bm{v}}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{J\bm{G}u},v}.
\end{align*}
and define the discrete physical gradient and divergence as
\begin{align*}
\LRp{\Grad_h u,v}_{\Omega} &= \LRp{J\bm{G}\widehat{\Grad}_h u,\bm{v}}_{\Omega} = \sum_{k}\LRp{J\bm{G}\widehat{\Grad}u,\bm{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{J\bm{G}\widehat{\bm{n}} \jump{u},\bm{v}} \\
\LRp{\Grad_h \cdot \bm{u},v}_{\Omega} &= \LRp{\widehat{\Grad}_h\cdot\LRp{J\bm{G}^T\bm{u}}}_{\Omega} = \sum_{k} \LRp{\widehat{\Grad}\cdot \LRp{J\bm{G}^T\bm{u}},{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}}. 
\end{align*}
The discrete gradient is consistent with the physical gradient.  The discrete divergence is also consistent with the physical divergence, provided that the jump term $\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}}$ is computed via
\[
\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}} = \LRa{\jump{J\LRp{\bm{G}\widehat{\bm{n}}}^T\bm{u}},v} = \LRa{\LRp{\widehat{\bm{n}}^T\LRp{J{\bm{G}}^T\bm{u}}}^+ + \LRp{\widehat{\bm{n}}^T\LRp{J{\bm{G}}^T\bm{u}}}^-,v}.  
\]

The advantage of this projected geometric factor approach is local conservation; taking $v = \bbm{1}_{D^k}$ recovers (after integration by parts) the flux integral
\[
\LRp{\Grad_h \cdot \bm{u},\bbm{1}_{D^k}}_{\Omega} = \int_{\partial D^k}\widehat{\bm{n}}^T\avg{J\bm{G}^T\bm{u}}.
\]
